The Form of Uncertainty Affects Selection for Social Learning

### 2023-02-22 ms.tex
more revisions

\paragraph{Payoff ambiguity.} Most models of social learning necessarily differentiate between the payoffs for adopting optimal vs. non-optimal behaviors
\citep{BoydRicherson1985,Rogers1988,Enquist2007,Rendell2010,aoki2014evolution}. The size of the difference between these payoffs is usually taken to influence the strength of selection on learning strategies.
However, the ability to discern payoff differences between behaviors is also a source of uncertainty. In reality, payoffs for particular behaviors are not always consistent. A behavior may yield a large payoff sometimes and a small
payoff other times \citep{McElreath2005}. This means that signals about the
relationship between behavior and payoff are often noisy, and differentiating
between behavioral options is in part a problem of signal detection. When the
difference in expected payoffs between optimal and non-optimal behaviors is very
large, this noise matters little, as the signal is still very clear. However, when
the expected payoffs of different behaviors are similar relative to the size of
their variances, ambiguity arises about which behaviors really are superior.
Smaller differences between payoffs corresponds to larger ambiguity. Payoff
ambiguity has been manipulated in both theoretical~\citep{perreault2012bayesian}
and empirical~\citep{McElreath2005,Toyokawa2019} studies, both of which support the
claim that payoff ambiguity increases the reliance on social information.
Importantly, payoff ambiguity %differs from selection-set size in that it not only 
affects both uncertainty \emph{and} the strength of selection, with smaller payoff differences leading to greater uncertainty for a learner and weaker evolutionary selection favoring optimal strategies. 
\edit{It is an outstanding problem to understand how payoff variance and payoff distributions generally affect evolved behavioral strategies~\citep{Haaland2019}. For the sake of theoretical clarity for our goal of understanding the \emph{interaction} of various sources of uncertainty we only operationalize payoff ambiguity as the difference between expected payoffs, even though empirical studies often include both forms of payoff uncertainty~\citep{McElreath2005,Toyokawa2019}, and a more complete model would include differences in mean and variance of payoffs.}

@@

% \begin{table}[h]
%     \caption{Outcome Variables. All averages are computed across trials at the end of the last generation.}
%     \label{tab:outcomeVariables}
%     \centering %\hspace{-3em}
%     \begin{tabular}{cp{4.25in}p{0.85in}} \toprule

%         Symbol & Description & Values \\

%         \midrule

%         $\meansl$ & Mean social-learning prevalence over agents
%                   & $ [0.0, 1.0]$ \\

%         $\meanpi / L$ & Mean payoffs accumulated across agents, normalized by
%         lifespan & $ [0.0, 1.0]$ \\

%         $\meanG$ & Mean number of generations to convergence & \{1,2,\ldots\} \\
%         \bottomrule
%     \end{tabular}
% \end{table}

@@

\clearpage
\begin{figure}
  \caption{Comparison of observed average payoffs in simulations with average payoffs
    obtained by populations homogeneously initialized to be all social or all asocial learners. Often the simulated payoffs follow the payoffs from the better-performing homogeneous group, with some exceptions discussed in the main text.}
  \label{fig:payoffs}
  \vspace{1em}
  \centering 
    {\textsf A, Low payoff $=$ 0.1, selection-set size $=$ 2} \\
    \begin{subfigure}[]{0.4\textwidth}
      \centering
      \includegraphics[width=\textwidth]{Figures/mean_prev_net_payoff_over_u_lowpayoff=0.1_nbehaviors=2.pdf}
    \end{subfigure} \\
    \centering
    {\textsf B, Low payoff $=$ 0.45, selection-set size $=$ 4} \\
    \begin{subfigure}[]{0.4\textwidth}
      % {\large \textsf B}
      \centering
      \includegraphics[width=\textwidth]{Figures/mean_prev_net_payoff_over_u_lowpayoff=0.45_nbehaviors=4.pdf}
    \end{subfigure} \\
    {\textsf C, Low payoff $=$ 0.45, selection-set size $=$ 10} \\
    \begin{subfigure}[]{0.4\textwidth}
      \centering
      % {\large \textsf C}
      \includegraphics[width=\textwidth]{Figures/mean_prev_net_payoff_over_u_lowpayoff=0.45_nbehaviors=10.pdf}
    \end{subfigure}

@@

We could have made alternative choices for a range of other structural features, the future exploration of which
will be important to the development of more parsimonious theories of social learning.  
For example, we only considered success-biased learning, although people may use conformity or other context biases to socially learn~\citep{BoydRicherson1985,Muthukrishna2016a,Smaldino2018b}. This modeling
choice maximizes the potential benefit to social learners, which suggests that
other mechanisms, like conformity, may lead to \edit{weaker selection} overall. 
\edit{Furthermore, our learning procedure could be interpreted as direct copying of
latent psychological variables (expected payoffs for each behavior); if children
learned instead from observations of the parent's behaviors without knowing rewards,
which in some cases would be a more realistic assumption~\citep{Wu2022b}, this may weaken the strength of selection for social learning.} We
also assumed that the number of behaviors and their payoffs were constant within a
given simulation, but this fails to account for evolutionary feedback which creates
new behavioral opportunities as time progresses\edit{~\citep{Chimento2022}}, e.g., via niche
construction~\citep{Smaldino2012a,Heras-Escribano2020} or cumulative cultural
evolution~\citep{Smolla2019,Derex2020}.  Group structure and processes such as
homophily and other group-level biases could inhibit the evolution of social
learning since successful out-group teachers could be cut off from in-group
learners~\citep{Golub2012}. We assumed agents choose behaviors via a softmax \edit{decision rule, which is more cognitively sophisticated than most social-learning models and captures the adaptiveness of real human learning mechanisms, which was essential for developing our understanding of how social and individual learning can combine to multiply problem solving capacity of our agents---however true human learning is significantly more sophisticated~\citep{Schulz2020a,Wu2022}}. More realistic and powerful individual learning could further support the evolution of
social learning. These considerations provide novel opportunities to expand the evolutionary theory of social learning by adding
such learning mechanisms to our model.

@@

%We found that the 
The evolution of social learning is a complex phenomenon, dependent on many interrelated factors, yet it is widespread across a
range of taxa. Although social learning is broadly studied, few studies have taken a systematic approach to understanding the role
of uncertainty in the evolution \edit{of} social learning. Understanding the role of uncertainty in social evolution is particularly important
for understanding human evolution as evidence increasingly suggests that the genus \emph{Homo} emerged in the context of substantial
(nonstationary) environmental variability \citep{anton_etal2014, levin2015}, and our global dispersal
suggests that human groups regularly face novel environments. \edit{In our model, more numerous behavioral options led agents to be more cognitively flexible via the softmax decision rule, and thus better able to adjust their behavior if social learning provided outdated information. While Rogers'-style models of social learning evolution may also predict greater social learning reliance in more challenging contexts (larger selection-set sizes), they cannot predict that social learning should be favored for such high values of
environmental variability as we observed here.} 
By carefully identifying and operationalizing different forms of uncertainty common in social-learning models, we developed a more thorough understanding of how uncertainty can favor social learning and how different forms of uncertainty interact to favor its evolution. By explicitly modeling a key individual-level learning mechanism shared across taxa, namely the ability to adjust behavior to uncertainty and new information, we found that social learning could evolve even if it provides outdated information. %Social learning is a critical component of problem solving among social creatures. Our work identified and systematized some important forms of uncertainty and catalogued their effects on the evolution of social learning.
Our work, then, both advances the theory of social learning and could eventually be of practical use in identifying specific mechanisms underlying adaptations to uncertain environments.

>>>>

\paragraph{Payoff ambiguity.} Most models of social learning necessarily differentiate between the payoffs for adopting optimal vs. non-optimal behaviors
\citep{BoydRicherson1985,Rogers1988,Enquist2007,Rendell2010,aoki2014evolution}. The size of the difference between these payoffs is usually taken to influence the strength of selection on learning strategies.
However, the ability to discern payoff differences between behaviors is also a source of uncertainty. In reality, payoffs for particular behaviors are not always consistent. A behavior may yield a large payoff sometimes and a small
payoff other times \citep{McElreath2005}. This means that signals about the
relationship between behavior and payoff are often noisy, and differentiating
between behavioral options is in part a problem of signal detection. When the
difference in expected payoffs between optimal and non-optimal behaviors is very
large, this noise matters little, as the signal is still very clear. However, when
the expected payoffs of different behaviors are similar relative to the size of
their variances, ambiguity arises about which behaviors really are superior.
Smaller differences between payoffs corresponds to larger ambiguity. Payoff
ambiguity has been manipulated in both theoretical~\citep{perreault2012bayesian}
and empirical~\citep{McElreath2005,Toyokawa2019} studies, both of which support the
claim that payoff ambiguity increases the reliance on social information.
Importantly, payoff ambiguity %differs from selection-set size in that it not only 
affects both uncertainty \emph{and} the strength of selection, with smaller payoff differences leading to greater uncertainty for a learner and weaker evolutionary selection favoring optimal strategies. 
\edit{It is an outstanding problem to understand how payoff variance and payoff
distributions generally affect evolved behavioral strategies~\citep{Haaland2019}.
For the sake of theoretical clarity for our goal of understanding the
\emph{interaction} of various sources of uncertainty we only operationalize payoff
ambiguity as the difference between expected payoffs, even though empirical studies
have sometimes included both forms of payoff uncertainty~\citep{McElreath2005,Toyokawa2019}, and a more complete model would include differences in mean and variance of payoffs.}

@@

\begin{table}[h]
    \caption{Outcome Variables. All averages are computed across trials at the end of the last generation.}
    \label{tab:outcomeVariables}
    \centering %\hspace{-3em}
    \begin{tabular}{clp{0.85in}} \toprule

        Symbol & Description & Values \\

        $\meansl$ & Mean social-learning prevalence over agents (Fig.~\ref{fig:mainResults})
                  & $ [0.0, 1.0]$ \\

        $\meanG$ & Mean number of generations to convergence (Fig.~\ref{fig:steps}) & \{1,2,\ldots\} \\

        $\meanpi / L$ & Mean payoffs accumulated across agents, normalized by
        lifespan (Fig.~\ref{fig:payoffs}) & $ [0.0, 1.0]$ \\

@@

\clearpage
\begin{figure}
  \caption{Comparison of observed average payoffs in simulations with average payoffs
    obtained by populations homogeneously initialized to be all social or all asocial learners. Often the simulated payoffs follow the payoffs from the better-performing homogeneous group, with some exceptions discussed in the main text.}
  \label{fig:payoffs}
  \vspace{1em}
  \centering 
    {\textsf A \quad Low payoff, $\pilow = 0.1$; selection-set size, $B = 2$} \\
    \begin{subfigure}[]{0.4\textwidth}
      \centering
      \includegraphics[width=\textwidth]{Figures/mean_prev_net_payoff_over_u_lowpayoff=0.1_nbehaviors=2.pdf}
      \\[-0.8em]
      {\centering \quad Environmental variability, $u$}
    \end{subfigure} \\[1.5em]
    \centering
    {\textsf B \quad Low payoff, $\pilow = 0.45$; selection-set size, $B = 4$} \\
    \begin{subfigure}[]{0.4\textwidth}
      % {\large \textsf B}
      \centering
      \includegraphics[width=\textwidth]{Figures/mean_prev_net_payoff_over_u_lowpayoff=0.45_nbehaviors=4.pdf}
      \\[-0.8em]
      {\centering \quad Environmental variability, $u$}
    \end{subfigure} \\[1.5em]
    {\textsf C \quad Low payoff, $\pilow = 0.45$; selection-set size, $B = 10$} \\
    \begin{subfigure}[]{0.4\textwidth}
      \centering
      % {\large \textsf C}
      \includegraphics[width=\textwidth]{Figures/mean_prev_net_payoff_over_u_lowpayoff=0.45_nbehaviors=10.pdf}
      \\[-0.8em]
      {\centering \quad Environmental variability, $u$}

    \end{subfigure}

@@

We could have made alternative choices for a range of other structural features, the future exploration of which
will be important to the development of more parsimonious theories of social learning.  
For example, we only considered success-biased learning, although people may use conformity or other context biases to socially learn~\citep{BoydRicherson1985,Muthukrishna2016a,Smaldino2018b}. This modeling
choice maximizes the potential benefit to social learners, which suggests that
other mechanisms, like conformity, may lead to \edit{weaker selection} overall. 
\edit{Furthermore, our learning procedure could be interpreted as direct copying of
latent psychological variables (expected payoffs for each behavior); if children
learned instead from observations of the parent's behaviors without knowing rewards,
which in some cases would be a more realistic assumption~\citep{Wu2022b}, this may weaken the strength of selection for social learning.} We
also assumed that the number of behaviors and their payoffs were constant within a
given simulation, but this fails to account for evolutionary feedback which creates
new behavioral opportunities as time progresses\edit{~\citep{Chimento2022}}, e.g., via niche
construction~\citep{Smaldino2012a,Heras-Escribano2020} or cumulative cultural
evolution~\citep{Smolla2019,Derex2020}.  Group structure and processes such as
homophily and other group-level biases could inhibit the evolution of social
learning since successful out-group teachers could be cut off from in-group
learners~\citep{Golub2012}. We assumed agents choose behaviors via a softmax
\edit{decision rule, which is more cognitively sophisticated than most
social-learning models and captures the adaptiveness of real human learning
mechanisms. This modeling decision was essential for developing our understanding of
how social and individual learning can combine to multiply problem solving capacity
of our agents, though it is still a pale shadow of the true sophistication of human
learning capabilities~\citep{Schulz2020a,Wu2022}}. More realistic and powerful individual learning could further support the evolution of
social learning. These considerations provide novel opportunities to expand the evolutionary theory of social learning by adding
such learning mechanisms to our model.

@@

%We found that the 
The evolution of social learning is a complex phenomenon, dependent on many interrelated factors, yet it is widespread across a
range of taxa. Although social learning is broadly studied, few studies have taken a systematic approach to understanding the role
of uncertainty in the evolution \edit{of} social learning. Understanding the role of uncertainty in social evolution is particularly important
for understanding human evolution as evidence increasingly suggests that the genus \emph{Homo} emerged in the context of substantial
(nonstationary) environmental variability \citep{anton_etal2014, levin2015}, and our global dispersal
suggests that human groups regularly face novel environments. \edit{In our model, more numerous behavioral options meant agents more flexibly adjusted their behavior if social learning provided outdated information since in the softmax decision rule, more options means relatively less weight is assigned to the most profitable behavior known to agents, and greater weight is put on unexplored behaviors or those behaviors perceived to be less optimal. While Rogers'-style models of social learning evolution may also predict greater social learning reliance in more challenging contexts (larger selection-set sizes), they cannot predict that social learning should be favored for such high values of
environmental variability as we observed here.} 
By carefully identifying and operationalizing different forms of uncertainty common in social-learning models, we developed a more thorough understanding of how uncertainty can favor social learning and how different forms of uncertainty interact to favor its evolution. By explicitly modeling a key individual-level learning mechanism shared across taxa, namely the ability to adjust behavior to uncertainty and new information, we found that social learning could evolve even if it provides outdated information. %Social learning is a critical component of problem solving among social creatures. Our work identified and systematized some important forms of uncertainty and catalogued their effects on the evolution of social learning.
Our work, then, both advances the theory of social learning and could eventually be of practical use in identifying specific mechanisms underlying adaptations to uncertain environments.
